---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# hdMTD

## Overview

hdMTD is a package designed for conducting inference in Mixture Transition Distribution (MTD) models, which can represent higher-order Markov models as a convex mixture of single-step Markov chains. The algorithms employed here are primarily built upon the work of [Ost and Takahashi (2023)](http://jmlr.org/papers/v24/22-0266.html) in their paper titled "Sparse Markov Models for High-Dimensional Inference."

## Installation

```{r, eval = FALSE}
remotes::install_github("MaiaraGripp/hdMTD")
```

## Usage

Given a sample from an MTD chain, the hdMTD() function estimates the relevant lag set $\Lambda$ using a specified method and a suitable set of parameters. The available methods include the *FSC* algorithm developed by [Ost and Takahashi (2023)](http://jmlr.org/papers/v24/22-0266.html), FSC's first step, referred to as *FS* method, its second step, referred to as *CUT* method, and a Bayesian Information Criterion (*BIC*) method.
Furthermore, the package can perform the following tasks:

* Create an object of class MTD with all the parameters necessary to define a proper MTD model using the `MTDmodel()` function. Such objects can be used to sample an MTD Markov Chain from its invariant distribution with the `perfectSample()` function.
* Calculate the oscillations of an MTD object or estimate them from a chain sample using the `oscillations()` function.
* Compute the Maximum Likelihood Estimates (MLE) of the transition probabilities for a chain sample given a relevant lag set with the `probs()` function.
* Determine the absolute frequency of each sequence of size $d$ that appears in a sample using the `countsTab()` function.
* Group such frequencies to consider only sequences time-indexed by a subset of ${-1, \dots, -d}$ with the `freqTabSj()` function.
* Lastly, `MTDest()` function uses a Expectation Maximization (EM) algorithm based on [Lèbre and Bourguinon (2008)](https://doi.org/10.1080/00949650701266666), to estimate the parameter set of an MTD model given a sample and a set of initial parameters.

```{r example}

library(hdMTD)

set.seed(1234)
Lambda <- c(1,5,10) #relevant lag set
A <- c(0,1) #state space
MTD <- MTDmodel(Lambda = Lambda,A=A);MTD # MTD object

oscillation(MTD)

X <- perfectSample(MTD=MTD,N=1000);X[1:10] # MTD chain sampled from invariant dist.

S1 <- hdMTD(X=X,d=15,method = "FS",l=3);S1 # estimate a relevant lag set of size 3 with the "Foward Stepwise" method.

S2 <- hdMTD(X,d=max(S1),method = "BIC",S=S1, minl=1, maxl=3);S2# estimate a relevant lag set with the "BIC" method using FS output.

S3 <- hdMTD(X,d=12,method = "BIC", minl=1, maxl=3, byl=TRUE, BICvalue=TRUE);S3
# estimate a relevant lag set with the "BIC" method.

S4 <- hdMTD(X,d=20,method = "CUT",S=S1);S4# estimate a relevant lag set with the "CUT" method using FS output.

p <- probs(X,S=S4, matrixform = TRUE);p #estimates the MLE transition probabilities given the CUT output.

init <- list(
  'lambdas'= c(0.05,0.3,0.3,0.35),
  'p0' = c(0.5,0.5),
  'pj' = list(
    matrix(c(0.5,0.5,
             0.5,0.5),ncol=2,nrow = 2),
    matrix(c(0.5,0.5,
             0.5,0.5),ncol=2,nrow = 2),
    matrix(c(0.5,0.5,
             0.5,0.5),ncol=2,nrow = 2)
  )
)

MTDest(X,S=c(1,5,10),init=init, iter = TRUE)
## Estimating P (transition probabilities)
estParam <- MTDest(X,S=c(1,15,30),init=init)
estMTD <- MTDmodel(Lambda,A,lam0=estParam$lambdas[1],
         lamj = estParam$lambdas[-1],p0=estParam$p0,
         pj=estParam$pj)
estMTD$P

```

## Data sets

This package includes three real-data sets that where acquired externaly. 

The raindata set was acquired at [Kaggle](https://www.kaggle.com/datasets/jsphyg/weather-dataset-rattle-package) but it's original source is Data source:[Australian Bureau of Meteorology] (http://www.bom.gov.au/climate/dwo/) and (http://www.bom.gov.au/climate/data). Copyright Commonwealth of Australia 2010, Bureau of Meteorology.

The sleepscoring data was acquired at the Haaglanden Medisch Centrum (HMC, The Netherlands) sleep center, and is available at Physionet.org. URL https://doi.org/10.13026/t79q-fr32.

The tempdata data set was acquired at INMET, the National Institute of Meteorology in Brazil https://bdmep.inmet.gov.br/.

## References

If you use our package in your work, please cite the following articles:

Ost G, Takahashi DY (2023). “Sparse Markov Models for High-dimensional Inference.” Journal of Machine Learning Research, 24(279), 1–54. 
URL http://jmlr.org/papers/v24/22-0266.html

Function MTDest was based on the ideas of the following article:

Lebre, Sophie & Bourguignon, Pierre-Yves. (2008). An EM algorithm for estimation in the Mixture Transition Distribution model. Journal of Statistical Computation and Simulation. 78. [Link to Article](https://doi.org/10.1080/00949650701266666)








